Notes on Learning Linear State Space Models

### 2015-08-30 notes/notes-linear-system-learning.tex
Changed code to F,Q formulation

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{subfig}
\usepackage{cite}
\usepackage[margin=1in]{geometry}

@@

There are two problems with extending the Gibbs sampling methodology to such a model. First, the state sequence and the transition covariance matrix are too strongly related. The sampled state sequence defines the subspace, so we cannot freely change $\lgtv$. Second, the conjugate prior is tricky. The matrix-normal-inverse-wishart can be generalised to the case where $\lgtv$ is not full rank (using singular wishart and singular matrix normal distributions). However, this is not helpful, because it constrains the transition matrix to lie in a linear subspace.

Because Wishart distributions are somewhat more studied than their inverse counterparts, it will be easier in the following section to work with the precision, defined as $\lgtp = (\lgdm\lgdm\tr)\pinv$. We will need the eigen-decomposition of this frequently.

\begin{IEEEeqnarray}{rCl}
 \lgtp &=& \tpvec \tpval \tpvec\tr
\end{IEEEeqnarray}
This is the ``non-singular'' factorisation, where $\tpval$ is a diagonal $\rk\times\rk$ matrix with $\rk = \rank(\lgtp)$, and $\tpvec$ from the appropriate Stiefel manifold. The full decomposition will also be useful at times.

\begin{IEEEeqnarray}{rCl}
 \lgtp &=& \begin{bmatrix}\tpvec & \tpvecorth \end{bmatrix} \begin{bmatrix}\tpval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tpvec\tr \\ \tpvecorth\tr \end{bmatrix}
\end{IEEEeqnarray}

@@

\subsubsection{Wishart Part}

There are two ways to extend the Wishart distribution to positive semi-definite matrices. Either let $\priordof \in \{z \in \naturals | z < \lsd\}$, in which case $\rank(\lgtp) = \priordof$, or let $\priorscalematrix$ be positive semi-definite. The latter makes it tricky to work out what we should set $\priorscalematrix$ to, so we'll go with the former. If we condition on a known value of the rank $r$, then the singular Wishart density is,

\begin{IEEEeqnarray}{rCl}
 \den(\lgtp) &=& \frac{ \pi^{\half\rk(\lsd-\rk)} \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tpval}^{\half(\rk-\lsd-1)} \exp\left( -\half \lgtp \priorscalematrix \right)    .
\end{IEEEeqnarray}

\cite{Uhlig1994,Diaz-Garcia1997}

@@

\begin{IEEEeqnarray}{rCl}
 \lgtm|\lgtp &\sim& \matrixnormaldist{\priormeanmatrix}{\lgtp\pinv+\tpvecorth\priorextratpval\inv\tpvecorth\tr}{\priorcolumnvariance}
\end{IEEEeqnarray}
%
$\priorextratpval$ is a diagonal matrix of positive eigenvalues which relax the constraint to lie in the subspace. They control the rate at which the density decays as we move away from that subspace. Setting these extra eigenvalues might be tricky. I haven't really worked out how to do this well yet. But they can depend on the other ones, so we could take the minimum, maximum or average of the others. Making them really big is uninformative.

@@

A sampled state trajectory $\ls{1:\timax}$ defines the subspace in which the state perturbations must lie, which means that neither $\lgtm$ nor $\lgtp$ can be freely changed, and Gibbs sampling alone does not work. However, we can still sample within the space specified by this constraint. Additional MCMC moves will then be needed which allow the subspace to be changed.

@@

Using Givens rotations, we can efficiently factorise the precision eigenvector matrix as follows.
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix} \tpvec & \tpvecorth \end{bmatrix} &=& \begin{bmatrix}\tprot & \tprotorth\end{bmatrix}\begin{bmatrix} \tpsign & \zmat \\ \zmat & \tpsignorth \end{bmatrix} \begin{bmatrix} \tprow & \zmat \\ \zmat & \idmat \end{bmatrix}\begin{bmatrix} \idmat & \zmat \\ \zmat & \tpnull \end{bmatrix}
\end{IEEEeqnarray}

So
%
\begin{IEEEeqnarray}{rCl}
 \tpvec     &=& \tprot     \tpsign     \tprow  \\
 \tpvecorth &=& \tprotorth \tpsignorth \tpnull
\end{IEEEeqnarray}

Then defined the following, which is positive definite
%
\begin{equation}
 \tpfull = \tpsign \tprow \tpval \tprow\tr \tpsign\tr
\end{equation}

So
%
\begin{IEEEeqnarray}{rCl}
 \lgtp &=& \tprot \tpfull \tprot\tr
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
 \ls{\ti} &=& \lgtm \ls{\ti-1} + \tprot \tpfull^{-\half} \tn{\ti} \\
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
 \Rightarrow \tprot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tpfull\inv}
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tpfull}^{\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \tpfull \tprot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)
\end{IEEEeqnarray}

The likelihood function associated with the state trajectory is:
\begin{align}
 \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
 &\propto \determ{\tpfull}^{\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \tpfull \tprot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
 &= \determ{\tpfull}^{\half(\timax-1)} \exp\left(-\half\trace\left[ \tpfull \tprot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tprot \right]\right) \nonumber \\
 &= \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \tprot\tr \lgtm \suffstats{1} \lgtm\tr \tprot - \tprot\tr \lgtm \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtm\tr \tprot + \tprot\tr \suffstats{3} \tprot \right) \right]\right) \\
 &= \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtmrot\tr + \tprot\tr \suffstats{3} \tprot \right) \right]\right)
\end{align}
%
where
%
\begin{equation}
 \lgtmrot = \tprot\tr \lgtm
\end{equation}

This is all subject to the constraint that each state must be reachable from the previous one, which we can write concisely as,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tprot Z
\end{IEEEeqnarray}
%
where $Z \in \reals^{\rk\times\lsd}$ and 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \begin{bmatrix} \ls{2} & \ls{3} & \hdots & \ls{\timax} \end{bmatrix}
\end{IEEEeqnarray}

@@

If we transform from $(\lgtm,\lgtp)$ to $(\lgtmrot,\tpfull)$, then the prior on these variables is as follows.

Using the transformation property of the matrix normal distribution:
%
\begin{IEEEeqnarray}{rCl}
 \lgtmrot|\lgtp = \lgtmrot|\tprot,\tpfull &\sim& \matrixnormaldist{\tprot\tr\priormeanmatrix}{\tpfull\inv}{\priorcolumnvariance}
\end{IEEEeqnarray}

Using the transformation property of the Wishart distribution:
%
\begin{IEEEeqnarray}{rCl}
 \tpfull|\tprot &\sim& \wishartdist{\rk}{\tprot\tr\priorscalematrix\tprot}
\end{IEEEeqnarray}

@@

We can now get to the conditional distribution for $\den(\lgtmrot, \tpfull|\tprot,\ls{1:\timax})$ which is well-behaved and can be sampled.

The parameter posterior conditional is:
\begin{equation}
 \den(\lgtmrot, \tpfull | \tprot, \ls{1:\timax}, \ob{1:\timax}) =  \den(\lgtmrot | \tpfull, \tprot, \ls{1:\timax}) \den(\tpfull | \tprot, \ls{1:\timax})
\end{equation}
\begin{align}
 \tpfull | \tprot, \ls{1:\timax} &\sim \wishartdist{\postdof}{\postscalematrix} \\
 \lgtmrot | \tpfull, \tprot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}
\end{align}

Hyperparameter updates:
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \tprot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &= \rk + \suffstats{0} \\
 \postscalematrix                        &= \tprot\tr\left( \priorscalematrix + \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tprot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{align}

Proof:
\begin{IEEEeqnarray}{rCl}
 \den(\lgtmrot, \tpfull|\tprot,\ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtmrot, \tpfull, \tprot) \den(\lgtmrot | \tpfull, \tprot) \den(\tpfull|\tprot) \nonumber \\
%
 &\propto& \determ{\tpfull}^{\half\suffstats{0}} \exp\left(-\half\trace\left[ \tpfull \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tprot - \tprot\tr \suffstats{2} \lgtmrot\tr + \tprot\tr \suffstats{3} \tprot \right) \right]\right) \nonumber \\
 & & \times \determ{\tpfull}^{\frac{r-r-1}{2}} \exp\left( -\half \trace\left[\tpfull \tprot\tr\priorscalematrix\tprot\right] \right) \nonumber \\
 & & \times \determ{\tpfull}^{\half} \exp\left(-\half \trace\left[ \tpfull (\lgtmrot-\tprot\tr\priormeanmatrix) \priorcolumnvariance\inv (\lgtmrot-\tprot\tr\priormeanmatrix)\tr \right] \right) \nonumber \\ 
%
 &=& \determ{\tpfull}^{\frac{(r+\suffstats{0})-r-1}{2}} \exp\left( -\half \trace\left[\tpfull \tprot\tr(\priorscalematrix+\suffstats{3})\tprot\right] \right) \nonumber \\
 & & \times \determ{\tpfull}^{\half} \exp\Bigg(-\half \trace\bigg[ \tpfull \bigg( \lgtmrot (\priorcolumnvariance\inv + \suffstats{1}) \lgtmrot\tr - \lgtmrot(\priorcolumnvariance\inv\priormeanmatrix\tr + \suffstats{2}\tr) \tprot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad  -\: \tprot\tr (\priormeanmatrix\priorcolumnvariance\inv + \suffstats{2}) \lgtmrot\tr + \tprot\tr \priormeanmatrix\priorcolumnvariance\inv\priormeanmatrix\tr \tprot \bigg) \bigg] \Bigg)
\end{IEEEeqnarray}
%
Finish by completing the square and comparing terms.

Of course, we want $\lgtm$, not $\lgtmrot$. We can recover the matrix uniquely using the constraint from the likelihood. The previous value of $\lgtm$ from before we sampled must have satisfied this constraint. Let's call that matrix $\lgtm\mcold$. 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm\mcold \LS{1:\timax-1} + \tprot Z\mcold \nonumber \\
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tprot Z \nonumber \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} + \tprot(Z-Z\mcold) \\
 \Rightarrow \zmat &=& (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} + (Z-Z\mcold) \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} - \tprot (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} \\
 \zmat &=& \left[(\lgtm-\lgtm\mcold) - \tprot (\lgtmrot-\lgtmrot\mcold)\right] \LS{1:\timax-1} \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) - \tprot (\lgtmrot-\lgtmrot\mcold)
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \lgtm\mcold + \tprot (\lgtmrot-\lgtmrot\mcold) \nonumber \\
       &=& \lgtm\mcold + \tprot (\lgtmrot-\tprot\tr\lgtm\mcold) \nonumber \\
       &=& (\idmat - \tprot\tprot\tr) \lgtm\mcold + \tprot \lgtmrot 
\end{IEEEeqnarray}

\subsection{Metropolis-Hastings for the Precision Subspace}

In order to change $\tprot$, we can target $\den(\lgtp|\lgtm, \ob{1:\timax})$ with Metropolis-Hastings, which constitutes a collapsed Gibbs move.

We sample a rotation matrix $\mhrot$ from some proposal distribution $\mhrotppsl$ and then apply the following transformation,
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix}
  \lgtp\mcnew \\ \mhrot\mcnew
 \end{bmatrix}
 & = &
 \begin{bmatrix}
  \mhrot\lgtp\mhrot\tr \\ \mhrot\tr
 \end{bmatrix} \nonumber     ,
\end{IEEEeqnarray}
%
which is clearly its own inverse. Since $\determ{\mhrot}=1$, it is straightforward to show that the Jacobian of this transformation is $1$, and hence using the reversible jump interpretation of Metropolis-Hastings \cite{Green1995,Green2009}, the acceptance probability is,
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtp\to\lgtp\mcnew) & = & \min\left\{1, \frac{ \den(\lgtp\mcnew|\lgtm,\ob{1:\timax})\mhrotppsl(\mhrot\mcnew) }{ \den(\lgtp|\lgtm, \ob{1:\timax})\mhrotppsl(\mhrot) } \right\}  \\
 & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtp\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtp)} \times \frac{\den(\lgtp\mcnew, \lgtm) }{\den(\lgtp, \lgtm)} \times \frac{\mhrotppsl(\mhrot\mcnew)}{\mhrotppsl(\mhrot)} \right\} \nonumber     .
\end{IEEEeqnarray}
%
The first term is simply a ratio of Kalman filter likelihoods.

@@

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{subfig}
\usepackage{cite}
\usepackage[margin=1in]{geometry}
\usepackage{color}

@@

There are two problems with extending the Gibbs sampling methodology to such a model. First, the state sequence and the transition covariance matrix are too strongly related. The sampled state sequence defines the subspace, so we cannot freely change $\lgtv$. Second, the conjugate prior is tricky. The matrix-normal-inverse-wishart can be generalised to the case where $\lgtv$ is not full rank (using singular inverse Wishart and singular matrix normal distributions). However, this is not helpful, because it constrains the transition matrix to lie in a linear subspace.

We will need the eigen-decomposition of $\lgtv$ frequently.

\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvvec \tvval \tvvec\tr
\end{IEEEeqnarray}
This is the ``non-singular'' part of the factorisation, where $\tvval$ is a diagonal $\rk\times\rk$ matrix with $\rk = \rank(\lgtv)$, and $\tvvec$ from the appropriate Stiefel manifold. The full decomposition will also be useful at times.

\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \begin{bmatrix}\tvvec & \tvvecorth \end{bmatrix} \begin{bmatrix}\tvval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tvvec\tr \\ \tvvecorth\tr \end{bmatrix}
\end{IEEEeqnarray}

@@

\subsubsection{Inverse Wishart Part}

There are two ways to extend the inverse Wishart distribution to positive semi-definite matrices. Either let $\priordof \in \{z \in \naturals | z < \lsd\}$, in which case $\rank(\lgtv) = \priordof$, or let $\priorscalematrix$ be positive semi-definite. The latter makes it tricky to work out what we should set $\priorscalematrix$ to, so we'll go with the former. If we set $\priordof$ equal to a known value of the rank $\rk$, then the singular Wishart density is,

\begin{IEEEeqnarray}{rCl}
 \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    .
\end{IEEEeqnarray}

\begin{itemize}
 \item \cite{Muirhead1982} covers the basic multivariate statistics for the full-rank case.
 \item \cite{Uhlig1994} covers singular Wishart distributions, including the eigenvalue distribution, and base measure for positive semi-definite matrices.
 \item \cite{Diaz-Garcia1997} covers singular and pseudo Wishart distributions in a thorough and unifying manner.
 \item \cite{Diaz-Garcia2006} covers singular and pseudo inverse Wishart distributions.
 \item \cite{Bodnar2008} contains a few useful properties.

\end{itemize}

@@

\begin{IEEEeqnarray}{rCl}
 \lgtm|\lgtv &\sim& \matrixnormaldist{\priormeanmatrix}{\lgtv+\tvvecorth\priorextratpval\tvvecorth\tr}{\priorcolumnvariance}
\end{IEEEeqnarray}
%
$\priorextratpval$ is a diagonal matrix of positive eigenvalues which relax the constraint to lie in the subspace. They control the rate at which the density decays as we move away from that subspace. Setting these extra eigenvalues might be tricky. I haven't really worked out how to do this well yet. But they can depend on the other ones, so we could take the minimum, maximum or average of the others. Making them really big is uninformative.

@@

A sampled state trajectory $\ls{1:\timax}$ defines the subspace in which the state perturbations must lie, which means that neither $\lgtm$ nor $\lgtv$ can be freely changed, and Gibbs sampling alone does not work. However, we can still sample within the space specified by this constraint. Additional MCMC moves will then be needed which allow the subspace to be changed.

@@

Using Givens rotations, we can efficiently factorise the precision eigenvector matrix as follows.
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix} \tvvec & \tvvecorth \end{bmatrix} &=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix}\begin{bmatrix} \tvsign & \zmat \\ \zmat & \tvsignorth \end{bmatrix} \begin{bmatrix} \tvrow & \zmat \\ \zmat & \idmat \end{bmatrix}\begin{bmatrix} \idmat & \zmat \\ \zmat & \tvnull \end{bmatrix}
\end{IEEEeqnarray}

So
%
\begin{IEEEeqnarray}{rCl}
 \tvvec     &=& \tvrot     \tvsign     \tvrow  \\
 \tvvecorth &=& \tvrotorth \tvsignorth \tvnull
\end{IEEEeqnarray}

Then defined the following, which is positive definite
%
\begin{equation}
 \tvfull = \tvsign \tvrow \tvval \tvrow\tr \tvsign\tr
\end{equation}

So
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr
\end{IEEEeqnarray}

@@

\begin{IEEEeqnarray}{rCl}
 \ls{\ti} &=& \lgtm \ls{\ti-1} + \tvrot \tvfull\msqrt \tn{\ti} \\
 \end{IEEEeqnarray}
 \begin{IEEEeqnarray}{rCl}
 \Rightarrow \tvrot\tr(\ls{\ti}-\lgtm\ls{\ti-1}) &\sim& \normaldist{\zmat}{\tvfull}
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) &\propto& \determ{\tvfull}^{-\half} \exp\left(-\half (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right)
\end{IEEEeqnarray}

The likelihood function associated with the state trajectory is:
\begin{align}
 \den(\ls{1:\timax}|\lgtm, \lgtv) &= \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) \nonumber \\
 &\propto \determ{\tvfull}^{-\half(\timax-1)} \exp\left( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \right) \nonumber \\
 &= \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &= \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right)
\end{align}
%
where
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm
\end{equation}

This is all subject to the constraint that each state must be reachable from the previous one, which we can write concisely as,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z
\end{IEEEeqnarray}
%
where $Z \in \reals^{\rk\times\lsd}$ and 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \begin{bmatrix} \ls{2} & \ls{3} & \hdots & \ls{\timax} \end{bmatrix}
\end{IEEEeqnarray}

@@

If we transform from $(\lgtm,\lgtv)$ to $(\lgtmrot,\tvfull)$, then the prior on these variables is as follows.

Using the transformation property of the matrix normal distribution:
%
\begin{IEEEeqnarray}{rCl}
 \lgtmrot|\lgtv = \lgtmrot|\tvrot,\tvfull &\sim& \matrixnormaldist{\tvrot\tr\priormeanmatrix}{\tvfull}{\priorcolumnvariance}
\end{IEEEeqnarray}

Using the transformation property of the Wishart distribution:
%
\begin{IEEEeqnarray}{rCl}
 \tvfull|\tvrot &\sim& \iwishartdist{\rk}{(\tvrot\tr\priorscalematrix\inv\tvrot)\inv}
\end{IEEEeqnarray}

@@

We can now get to the conditional distribution for $\den(\lgtmrot, \tvfull|\tvrot,\ls{1:\timax})$ which is well-behaved and can be sampled.

The parameter posterior conditional is:
\begin{equation}
 \den(\lgtmrot, \tvfull | \tvrot, \ls{1:\timax}, \ob{1:\timax}) =  \den(\lgtmrot | \tvfull, \tvrot, \ls{1:\timax}) \den(\tvfull | \tvrot, \ls{1:\timax})
\end{equation}
\begin{align}
 \tvfull | \tvrot, \ls{1:\timax} &\sim \iwishartdist{\postdof}{\postscalematrix} \\
 \lgtmrot | \tvfull, \tvrot, \ls{1:\timax} &\sim \matrixnormaldist{\postmeanmatrix}{\lgtv}{\postcolumnvariance}
\end{align}

Hyperparameter updates:
\begin{align}
 \postcolumnvariance\inv                 &= \priorcolumnvariance\inv + \suffstats{1} \\
 \postmeanmatrix \postcolumnvariance\inv &= \tvrot\tr\left( \priormeanmatrix \priorcolumnvariance\inv + \suffstats{2} \right) \\
 \postdof                                &= \rk + \suffstats{0} \\
 \postscalematrix                        &= (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\left( \suffstats{3} + \priormeanmatrix \priorcolumnvariance\inv \priormeanmatrix\tr \right)\tvrot - \postmeanmatrix \postcolumnvariance\inv \postmeanmatrix\tr
\end{align}

Proof:
\begin{IEEEeqnarray}{rCl}
 \den(\lgtmrot, \tvfull|\tvrot,\ls{1:\timax}) &\propto& \den(\ls{1:\timax}|\lgtmrot, \tvfull, \tvrot) \den(\lgtmrot | \tvfull, \tvrot) \den(\tvfull|\tvrot) \nonumber \\
%
 &\propto& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \nonumber \\
 & & \times \determ{\tvfull}^{-\half(r+r+1)} \exp\left( -\half \trace\left[\tvfull\inv (\tvrot\tr\priorscalematrix\inv\tvrot)\inv \right] \right) \nonumber \\
 & & \times \determ{\tvfull}^{-\half} \exp\left(-\half \trace\left[ \tvfull\inv (\lgtmrot-\tvrot\tr\priormeanmatrix) \priorcolumnvariance\inv (\lgtmrot-\tvrot\tr\priormeanmatrix)\tr \right] \right) \nonumber \\ 
%
 &=& \determ{\tvfull}^{-\half\left((r+\suffstats{0})+r+1\right)} \exp\left( -\half \trace\left[\tvfull\inv \left( (\tvrot\tr\priorscalematrix\inv\tvrot)\inv + \tvrot\tr\suffstats{3}\tvrot \right)\right] \right) \nonumber \\
 & & \times \determ{\tvfull}^{-\half} \exp\Bigg(-\half \trace\bigg[ \tvfull\inv \bigg( \lgtmrot (\priorcolumnvariance\inv + \suffstats{1}) \lgtmrot\tr - \lgtmrot(\priorcolumnvariance\inv\priormeanmatrix\tr + \suffstats{2}\tr) \tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad  -\: \tvrot\tr (\priormeanmatrix\priorcolumnvariance\inv + \suffstats{2}) \lgtmrot\tr + \tvrot\tr \priormeanmatrix\priorcolumnvariance\inv\priormeanmatrix\tr \tvrot \bigg) \bigg] \Bigg)
\end{IEEEeqnarray}
%
Finish by completing the square and comparing terms.

Of course, we want $\lgtm$, not $\lgtmrot$. We can recover the matrix uniquely using the constraint from the likelihood. The previous value of $\lgtm$ from before we sampled must have satisfied this constraint. Let's call that matrix $\lgtm\mcold$. 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm\mcold \LS{1:\timax-1} + \tvrot Z\mcold \nonumber \\
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z \nonumber \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} + \tvrot(Z-Z\mcold) \\
 \Rightarrow \zmat &=& (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} + (Z-Z\mcold) \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) \LS{1:\timax-1} - \tvrot (\lgtmrot-\lgtmrot\mcold) \LS{1:\timax-1} \\
 \zmat &=& \left[(\lgtm-\lgtm\mcold) - \tvrot (\lgtmrot-\lgtmrot\mcold)\right] \LS{1:\timax-1} \\
 \Rightarrow \zmat &=& (\lgtm-\lgtm\mcold) - \tvrot (\lgtmrot-\lgtmrot\mcold)
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \lgtm\mcold + \tvrot (\lgtmrot-\lgtmrot\mcold) \nonumber \\
       &=& \lgtm\mcold + \tvrot (\lgtmrot-\tvrot\tr\lgtm\mcold) \nonumber \\
       &=& (\idmat - \tvrot\tvrot\tr) \lgtm\mcold + \tvrot \lgtmrot 
\end{IEEEeqnarray}

\subsubsection{Metropolis-Hastings for the Covariance Subspace}

In order to change $\tvrot$, we can target $\den(\lgtv|\lgtm, \ob{1:\timax})$ with Metropolis-Hastings, which is justified as a collapsed Gibbs move.

We sample a rotation matrix $\mhrot$ from some proposal distribution $\mhrotppsl$ and then apply the following transformation,
%
\begin{IEEEeqnarray}{rCl}
 \begin{bmatrix}
  \lgtv\mcnew \\ \mhrot\mcnew
 \end{bmatrix}
 & = &
 \begin{bmatrix}
  \mhrot\lgtv\mhrot\tr \\ \mhrot\tr
 \end{bmatrix} \nonumber     ,
\end{IEEEeqnarray}
%
which is clearly its own inverse. Since $\determ{\mhrot}=1$, it is straightforward to show that the Jacobian of this transformation is $1$, and hence using the reversible jump interpretation of Metropolis-Hastings \cite{Green1995,Green2009}, the acceptance probability is,
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtv\to\lgtv\mcnew) & = & \min\left\{1, \frac{ \den(\lgtv\mcnew|\lgtm,\ob{1:\timax})\mhrotppsl(\mhrot\mcnew) }{ \den(\lgtv|\lgtm, \ob{1:\timax})\mhrotppsl(\mhrot) } \right\}  \\
 & = & \min\left\{1, \frac{\den(\ob{1:\timax}|\lgtm,\lgtv\mcnew)}{\den(\ob{1:\timax}|\lgtm,\lgtv)} \times \frac{\den(\lgtv\mcnew, \lgtm) }{\den(\lgtv, \lgtm)} \times \frac{\mhrotppsl(\mhrot\mcnew)}{\mhrotppsl(\mhrot)} \right\} \nonumber     .
\end{IEEEeqnarray}
%
The first term is simply a ratio of Kalman filter likelihoods.

@@

\subsubsection{Metropolis-Hastings for the Transition Matrix}

Define $\postden$ as the posterior density, conditioning on $\ob{1:\timax}$.

If we modify the covariance matrix slightly so that it is no longer singular, then we can use the existing sampling procedure from the basic linear model and treat the resulting value of $\lgtm$ as a proposal in a Metropolis-Hastings kernel targeting $\postden(\lgtm, \ls{1:\timax}|\lgtv)$.
